# Week 6 - Analysis of longitudinal data

This first chunk is just for libraries.
```{r}

date()

library(corrplot)
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)

```

## Task 1. Analysing RATS data

First, let's load RATS data set and look its structure

```{r}

url_RATS="https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt"

RATS=read.csv(url_RATS, sep="\t")

RATS$ID=as.factor(RATS$ID)
RATS$Group=as.factor(RATS$Group)


glimpse(RATS)
head(RATS)
tail(RATS)

```

The RATS file has 16 time series with 11 time series each. The time serieses can be spitted ont three different groups.


Now, let's convert the RATS into "long format"

```{r}

RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 

str(RATSL)

glimpse(RATSL)
head(RATSL)
tail(RATSL)
```

Let's draw each individual time series.


```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, group = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```

It can be seen, that group 1 is clearly separated data set. Group 2 has one clear out lier data series. Groups 2 and 3 seems not to differ much.

Let's standardize the data sets. And then plot them.


```{r}
# Standardise the scores:
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
p1 <- ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized Weight")
p6
```

When standardized the limes looks more or less the same as without standardization.

Let's then look the box plots.


```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.8,0.8))
p5 <- p4 + scale_x_discrete(name = "week")
# Black & White version:
#p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p5
```


With the box plots, the difference between groups 2 and 3 comes clear. However, each of the groups have clearsly outliers.

Let's look the mean values for each group.

```{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight)")
p5
```

The most obvious outlier is for Group 2. Let's remove it.

```{r}
# Remove the outlier:
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
p1 <- ggplot(RATSL8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), weeks 1-8")
p5
```

Now, let's test, how good linear fitting we can do?

```{r}
# Add the baseline from the original data as a new variable to the summary data:
baseline <- RATS$WD1
RATSL8S2 <- RATSL8S %>%
  mutate(baseline)
glimpse(RATSL8S2)
# Fit the ANCOVA model and see the results:
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)
anova(fit)
```

It seems that we can separate the different groups even when the baseline is taken into account.

## Task 2. Analyzing the BPRS Data

First, let's load the dataset and modify it a bit.

```{r}

url="https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt"

BPRS=read.csv(url, sep=" ")

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)

# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))

```

The dataset BRPS consists 40 rows and 11 variables. Two of the variables are related to identifying and nine are the actual time series.

```{r}

glimpse(BPRSL)

```



```{r}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ treatment, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
p6
```



```{r, echo=TRUE}
BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRSL_reg)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```



```{r, echo=TRUE}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "Weight (grams)")
p5 <- p4 + theme_bw() + theme(legend.position = "top")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

```{r, echo=TRUE, fig.width=10, fig.height=10}
pairs(BPRS, cex = 0.7)
```

```{r, echo=TRUE}
library("lme4")
BPRSL_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRSL_ref)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```

```{r, echo=TRUE}
BPRSL_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRSL_ref1)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
anova(BPRSL_ref1, BPRSL_ref)
```


```{r, echo=TRUE}
BPRSL_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRSL_ref2)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
anova(BPRSL_ref1, BPRSL_ref2)
```



```{r, echo=TRUE, fig.width=3, fig.height=4}
Fitted <- fitted(BPRSL_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "Weight (grams)")
p5 <- p4 + theme_bw() + theme(legend.position = "right") # "none" in the book
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7

p1 <- ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "Weight (grams)")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7
graph1; graph2
```

